---
title: "Final Project"
subtitle: "Inferential Statistics 2025/2026"
author: "Giordano Alberti, Leonardo Bertani"
date: "`r Sys.Date()`"
output: 
  pdf_document: 
    toc: true
    fig_height: 6
    number_sections: true
    extra_dependencies: ["longtable", "booktabs", "array"]
  html_document: 
    toc: true
editor_options: 
  markdown: 
    wrap: 80
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\newpage

# Abstract

This report analyzes the determinants of academic performance using the UCI "Student Performance Data Set," derived from Mathematics classes in two Portuguese schools (Gabriel Pereira and Mousinho da Silveira) and collected through school reports and questionnaires. The study aims to identify the demographic, social, and behavioral factors driving final grades (G3), using empirical evidence to validate or refuse common assumptions about student performance. Employing Chi-Squared tests, along with parametric tests and regression analysis, we investigate variables ranging from family background to social habits.

# Introduction


# Data description and exploration

```{r include=FALSE}
data <- read.csv("./data/student-mat.csv", sep=";")
str(data)
summary(data)
head(data)
```

# Statistical Analysis

## Linear Regression (Numerical Prediction)

We aim to predict the final grade (G3) using linear regression models. This allows us to quantify the impact of specific variables while controlling for others.

### Model A: "Effort"

This model attempts to explain performance based on student effort and academic history.
$$ G3 = \beta_0 + \beta_1 \cdot \text{studytime} + \beta_2 \cdot \text{absences} + \beta_3 \cdot \text{failures} + \epsilon $$

```{r}
model_a <- lm(G3 ~ studytime + absences + failures, data = data)
summary(model_a)
```

**Interpretation:**
The model explains approximately **13.5%** of the variance in grades ($R^2 \approx 0.135$).
*   **Failures**: This is the most significant predictor ($t \approx -7.47$). Each past failure is associated with a decrease of about **2.2 points** in the final grade.
*   **Studytime**: Surprisingly, this variable is not statistically significant in this specific model ($p > 0.05$), suggesting that simply accounting for hours studied (without considering quality or efficiency) does not strongly predict G3 when checking for past failures.
*   **Absences**: Also not significant in this combination.

### Model B: "Background"

This model focuses on family and social environment.
$$ G3 = \beta_0 + \beta_1 \cdot \text{Medu} + \beta_2 \cdot \text{famrel} + \beta_3 \cdot \text{internet} + \epsilon $$

```{r}
model_b <- lm(G3 ~ Medu + famrel + internet, data = data)
summary(model_b)
```

**Interpretation:**
This model has a lower explanatory power ($R^2 \approx 0.05$).
*   **Medu (Mother's Education)**: This is a significant positive predictor ($t \approx 4.10$). Higher maternal education correlates with higher grades (approx. +0.86 points per level).
*   **Famrel & Internet**: Neither family relationship quality nor internet access are statistically significant predictors in this model when Mother's Education is included.

### Step-wise Selection

To find the most parsimonious model that explains the data best, we use a stepwise selection procedure (based on AIC) starting from a model including all the variables analyzed above.

```{r}
# Full model with candidates from A and B
full_model <- lm(G3 ~ studytime + absences + failures + Medu + famrel + internet, data = data)
best_model <- step(full_model, direction = "both", trace = 0)
summary(best_model)
```

The stepwise procedure generally converges to a model retaining **Failures** and **Medu** as the core predictors. This reinforces that past academic history (failures) and socio-economic background (mother's education) are the strongest available drivers for the final grade in this dataset, outweighing self-reported study time or social habits.
