---
title: "Final Project"
subtitle: "Inferential Statistics 2025/2026"
author: "Giordano Alberti, Leonardo Bertani"
date: "`r Sys.Date()`"
output: 
  pdf_document: 
    toc: true
    fig_height: 6
    number_sections: true
    extra_dependencies: ["longtable", "booktabs", "array"]
    toc_depth: 3
  html_document: 
    toc: true
editor_options: 
  markdown: 
    wrap: 80
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\newpage

# Abstract



# Introduction


# Data description and exploration

```{r include=FALSE}
data <- read.csv("./data/student-mat.csv", sep=";")
str(data)
summary(data)
head(data)
```

# Statistical Analysis

## T-Test (Group Comparison)

The Student's t-test is a parametric statistical method used to determine if there is a significant difference between the means of two independent groups. It assumes that the data follows a normal distribution and that the observations are independent.

We test the following hypotheses:

*   $H_0: \mu_1 = \mu_2$ (There is no difference in mean grades between the groups)
*   $H_1: \mu_1 \neq \mu_2$ (There is a significant difference in mean grades)

The test statistic $t$ is calculated based on the difference in sample means relative to the standard error of the difference. If the resulting p-value is less than our significance level ($\alpha = 0.05$), we reject the null hypothesis.

### Hypothesis A (Technological)

Is there a significant difference in grades between those who have Internet at home and those who do not?

```{r, echo=FALSE, include=FALSE}
t.test(G3 ~ internet, data = data)
```

The results indicate that students with Internet access at home have a higher mean grade compared to those without. The T-test yields a p-value slightly below the 0.05 threshold or right on the margin depending on the variance assumption, but generally suggests a positive tendency. Specifically, the mean grade for students with Internet is approximately **10.6**, while for those without, it is **9.4**. 

This difference might stem from easier access to study materials, online exercises, and educational platforms. However, since the access to the internet nowadays is ubiquitous, the "no" group might represent a specific socio-economic demographic that faces other challenges affecting their performance.

### Hypothesis B (Sentimental)

Do students in a romantic relationship have lower grades compared to single students?

```{r, echo=FALSE, include=FALSE}
t.test(G3 ~ romantic, data = data)
```

Here we observe a statistically significant difference. The mean grade for single students is approximately **10.8**, whereas students in a romantic relationship have a mean grade of **9.6**. The p-value confirms that this difference is unlikely to be due to chance.

This finding aligns with the "time allocation" theory: maintaining a romantic relationship requires time and emotional energy that might otherwise be dedicated to studying. The observed drop of more than 1 grade point is a substantial impact on the final performance.

## ANOVA (Group Comparison: Multiple Groups)

Analysis of Variance (ANOVA) is a statistical method used to test differences between two or more means. It generalizes the t-test to more than two groups by analyzing the variance *between* groups relative to the variance *within* groups.

We test the following hypotheses:

*   $H_0: \mu_1 = \mu_2 = \dots = \mu_k$ (All group means are equal)
*   $H_1$: At least one group mean is different from the others

The F-statistic determines the ratio of explained variance (between groups) to unexplained variance (within groups). A significant p-value ($< 0.05$) indicates that the factor has a statistically significant effect on the final grade.

### Hypothesis A (Social/Classical)

Does the mother's occupation influence the child's academic performance?

```{r, echo=FALSE, include=FALSE}
summary(aov(G3 ~ Mjob, data = data))
```

The results show significant differences in mean grades across different mother's occupations (F = 3.75, p < 0.05). Specifically:

- **Health professionals**: Mean G3 = 12.15 (highest)
- **Teachers**: Mean G3 = 11.05
- **Services**: Mean G3 = 11.02
- **Other**: Mean G3 = 9.82
- **At home**: Mean G3 = 9.15 (lowest)

The analysis suggests that having a mother in the health or education sector is associated with higher student performance. This could reflect:

1. Higher educational attainment of such parents, leading to better support at home
2. Greater awareness of the importance of education
3. Potentially better socio-economic conditions

In contrast, mothers who work "at home" or in "other" professions show children with lower grades, though this may also correlate with socio-economic factors not directly captured.

### Hypothesis B (Behavioral)

Does weekend alcohol consumption negatively impact grades?

```{r, echo=FALSE, include=FALSE}
summary(aov(G3 ~ as.factor(Walc), data = data))
```

Interestingly, the results show **no significant difference** in grades across weekend alcohol consumption levels (F = 0.73, p > 0.05). The mean grades are relatively consistent:

- Level 1 (very low): Mean G3 = 10.74
- Level 2 (low): Mean G3 = 10.08
- Level 3 (medium): Mean G3 = 10.73
- Level 4 (high): Mean G3 = 9.69
- Level 5 (very high): Mean G3 = 10.14

While there is a slight downward trend at level 4, the overall variation is not statistically significant. This suggests that weekend alcohol consumption alone, without considering frequency or context, may not be a primary driver of academic performance differences in this dataset. However, it's worth noting that social context and peer dynamics (which may be correlated with alcohol use) could have confounding effects.

### Hypothesis C (Travel/Logistical)

Does travel time from home to school affect academic performance?

```{r, echo=FALSE, include=FALSE}
summary(aov(G3 ~ traveltime, data = data))
```

The results show a marginally significant effect of travel time (F = 1.84, p $\approx$ 0.14), with a clear downward trend:

- **1 hour or less**: Mean G3 = 10.78 (n=257)
- **1-2 hours**: Mean G3 = 9.91 (n=107)
- **2-3 hours**: Mean G3 = 9.26 (n=23)
- **>3 hours**: Mean G3 = 8.75 (n=8)

The data suggests a practical relationship: longer commutes are associated with lower grades. This could be due to fatigue, reduced study time after accounting for commute, or the stress of long travel. The effect becomes increasingly pronounced at the extremes (>3 hours), though the sample size is very small in this group.

## Linear Regression (Numerical Prediction)

Linear Regression is a statistical method used to model the relationship between a dependent variable $Y$ and one or more independent variables $X$. The goal is to find the linear equation that best predicts the value of the dependent variable based on the predictors.

The general formula is:
$$ Y = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \dots + \beta_k X_k + \epsilon $$
Where:

  -   $\beta_0$ is the intercept.
  -   $\beta_i$ are the coefficients representing the change in $Y$ for a one-unit change in $X_i$.
  -   $\epsilon$ is the error term (residuals).

We aim to predict the final grade (**G3**) based on different sets of predictors.

### Pre-analysis: Correlation Check

Before building the models, it is crucial to examine the relationships between all candidate numeric variables to check for multicollinearity, which could distort valid statistical inferences. We examine the correlation between potential predictors including age, parental education, behavioral factors, and health.

```{r, echo=FALSE}
library(corrplot)
numeric_columns <- c("age", "Medu", "Fedu", "traveltime", "studytime", "failures", 
                     "famrel", "freetime", "goout", "Dalc", "Walc", "health", "absences")
numeric_vars <- data[, numeric_columns]
cor_matrix <- cor(numeric_vars, use = "complete.obs")
corrplot(cor_matrix, method = "color", type = "upper", tl.cex = 0.7, addCoef.col = "black", number.cex = 0.6)
```

The correlation matrix reveals two notable clusters of collinearity:

1.  **Parental Education**: `Medu` and `Fedu` have a strong positive correlation ($r \approx 0.62$). Since they carry similar information (socioeconomic status), including both might inflate standard errors. We will primarily use `Medu` as it often has a slightly stronger predictive power in educational contexts.
2.  **Alcohol Consumption**: `Dalc` (workday) and `Walc` (weekend) are strongly correlated ($r \approx 0.65$). We might consider combining them or selecting one.

Other correlations are generally moderate to weak ($|r| < 0.35$), suggesting that aside from these specific pairs, multicollinearity is manageable. The strongest negative correlation remains between `Medu` and `failures` ($\approx -0.24$).

### Model A: "Academic Effort"

This model focuses on student behavior and academic history. The hypothesis is that effort (`studytime`), attendance (`absences`), and past performance (`failures`) are the main drivers of the final grade.

$$ G3 = \beta_0 + \beta_1 \cdot \text{studytime} + \beta_2 \cdot \text{absences} + \beta_3 \cdot \text{failures} + \epsilon $$

```{r, echo=FALSE}
model_a <- lm(G3 ~ studytime + absences + failures, data = data)
summary(model_a)
```

**Interpretation:**

*   **Model Fit**: The $R^2$ is **0.135**, meaning this model explains about 13.5% of the variability in the final grade.
*   **Significance**:
    *   **Failures ($\beta \approx -2.20$)**: Highly significant ($p < 0.001$). Each past failure reduces the expected final grade by over 2 points.
    *   **Studytime**: Not significant ($p = 0.41$). Surprisingly, self-reported study time does not directly predict G3 when controlling for failures.
    *   **Absences**: Not significant.

### Model B: "Background & Environment"

This model tests the "Nature/Nurture" hypothesis: that family background and resources determine success.

$$ G3 = \beta_0 + \beta_1 \cdot \text{Medu} + \beta_2 \cdot \text{famrel} + \beta_3 \cdot \text{internet} + \epsilon $$

```{r, echo=FALSE}
model_b <- lm(G3 ~ Medu + famrel + internet, data = data)
summary(model_b)
```

**Interpretation:**

*   **Model Fit**: The $R^2$ is only **0.053** (5.3%). This model performs worse than the "Effort" model.
*   **Significance**:
    *   **Medu ($\beta \approx 0.86$)**: Significant ($p < 0.001$). Mother's education is a strong proxy for student success.
    *   **Internet/Famrel**: Neither are statistically significant predictors in this specific combination.

### Step-wise Selection (Optimization)

To find the most parsimonious model (the "Best Fit"), we use a stepwise selection procedure. This algorithm iteratively adds or removes variables to balance model fit against complexity. We start with a **Full Model** containing all potential predictors (excluding the intermediate grades G1 and G2 to focus on socio-demographic drivers).

```{r, echo=FALSE}
# Full model with all variables except G1 and G2
full_model <- lm(G3 ~ . - G1 - G2, data = data)
best_model <- step(full_model, direction = "both", trace = 0)

# Extract coefficients
coefs_table <- summary(best_model)$coefficients

# Filter to show only significant predictors (p < 0.05) 
# and remove the Intercept for a cleaner view of drivers
significant_preds <- coefs_table[coefs_table[, "Pr(>|t|)"] < 0.05 & rownames(coefs_table) != "(Intercept)", ]

# formatting the table
knitr::kable(significant_preds[, c("Estimate", "Pr(>|t|)")], 
             digits = 4, 
             caption = "Significant Predictors Selected by Stepwise Model")
```

The stepwise algorithm selects a model that includes **Failures**, **Mother's Job**, **Family Relationship**, **Free time**, **Absences**, and **Mother's Education**.
This confirms that academic history (**failures**) is the strongest driver, but family background (**mother's job/education**) and behavior (**absences/go out**) also play significant roles.

### Model Diagnostics: The "Zero" Problem

To validate our "Best Fit" model, we perform diagnostic checks on the residuals.

#### Normality of Residuals (Q-Q Plot)

The linear regression model assumes that the residuals (errors) are normally distributed.

```{r fig.height=5, fig.width=6, fig.align="center", echo=FALSE}
library(ggplot2)

# Extract residuals
residuals_df <- data.frame(residuals = residuals(best_model))

# Create Q-Q Plot with ggplot2
ggplot(residuals_df, aes(sample = residuals)) +
  geom_qq(color = "steelblue", alpha = 0.6) +
  geom_qq_line(color = "red", linetype = "dashed", linewidth = 1) +
  labs(title = "Normal Q-Q Plot (Full Data)",
       subtitle = "Notice the deviation in the lower tail",
       x = "Theoretical Quantiles",
       y = "Sample Quantiles") +
  theme_minimal()
```

**Interpretation:**
Most points follow the line, but there is a **significant deviation in the lower tail**. This indicates that the residuals are not perfectly normal. This is primarily due to the presence of students with a grade of **0**. A grade of 0 in this dataset is often a symbolic value indicating data about an **administrative failure** (dropout or missed exam), rather than an **academic failure** (insufficient grade between 1 and 9).

Our model consistently **overestimates** these grades (predicting e.g., a 8 or 9 when the actual is 0), leading to large negative residuals.

#### Actual vs Predicted

```{r fig.height=5, fig.width=6, fig.align="center", echo=FALSE}
# Create a dataframe with predictions
predictions <- predict(best_model)
results <- data.frame(Actual = data$G3, Predicted = predictions)

# Plot Actual vs Predicted with Jitter to show overlapping points
ggplot(results, aes(x = Actual, y = Predicted)) +
  geom_jitter(color = "steelblue", alpha = 0.6, width = 0.2, height = 0.2) +
  geom_abline(intercept = 0, slope = 1, color = "darkred", linetype = "dashed", linewidth = 1) +
  labs(title = "Actual vs Predicted Grades (Full Data)",
       subtitle = "Model struggles with the 0-grade cluster",
       x = "Actual Grade",
       y = "Predicted Grade") +
  coord_fixed(ratio = 1, xlim = c(0, 20), ylim = c(0, 20)) +
  theme_minimal()
```

The plot clearly shows the cluster of points at $Actual=0$ where predictions are much higher. This confirms that "administrative failure" (0) is a distinct event from "academic failure" (low grade), and treating them as a continuous scale introduces noise.

### Refining the Model: Assessing Non-Zero Grades

Since we identified that 0 grades are structural anomalies (dropouts), we refine our Linear Regression model to focus only on active students who received a grade ($G3 > 0$). We retrain the same model structure on this subset to assess its real predictive power for academic performance.

```{r, echo=FALSE}
# Filter for non-zero grades
data_nonzero <- subset(data, G3 > 0)

# Retrain the "best_model" structure on the filtered data
refined_model <- update(best_model, data = data_nonzero)
# summary(refined_model)

# Extract coefficients
coefs_table <- summary(refined_model)$coefficients
# Filter to show only significant predictors (p < 0.05) 
# and remove the Intercept for a cleaner view of drivers
significant_preds <- coefs_table[coefs_table[, "Pr(>|t|)"] < 0.05 & rownames(coefs_table) != "(Intercept)", ]

# formatting the table
knitr::kable(significant_preds[, c("Estimate", "Pr(>|t|)")], 
             digits = 4, 
             caption = "Significant Predictors Selected by Stepwise Model")
```

**Diagnostics of Refined Model:**

```{r fig.height=5, fig.width=6, fig.align="center", echo=FALSE}
residuals_refined <- data.frame(residuals = residuals(refined_model))

ggplot(residuals_refined, aes(sample = residuals)) +
  geom_qq(color = "forestgreen", alpha = 0.6) +
  geom_qq_line(color = "red", linetype = "dashed", linewidth = 1) +
  labs(title = "Normal Q-Q Plot (G3 > 0 Only)",
       subtitle = "Normality assumption is much better satisfied",
       x = "Theoretical Quantiles",
       y = "Sample Quantiles") +
  theme_minimal()
```

We also visualize the prediction accuracy for this refined subset:

```{r fig.height=5, fig.width=6, fig.align="center", echo=FALSE}
# Predictions for the refined model
pred_refined <- predict(refined_model)
results_refined <- data.frame(Actual = data_nonzero$G3, Predicted = pred_refined)

ggplot(results_refined, aes(x = Actual, y = Predicted)) +
  geom_jitter(color = "forestgreen", alpha = 0.6, width = 0.2, height = 0.2) +
  geom_abline(intercept = 0, slope = 1, color = "darkred", linetype = "dashed", linewidth = 1) +
  labs(title = "Actual vs Predicted Grades (Refined Model)",
       subtitle = "Prediction pattern is closer to the ideal line",
       x = "Actual Grade (G3 > 0)",
       y = "Predicted Grade") +
  coord_fixed(ratio = 1, xlim = c(0, 20), ylim = c(0, 20)) +
  theme_minimal()
```

We observe the following improvements:

1.  **Model Validity**: The refined model's $R^2$ increased to **0.275** (from ~0.15), and the Q-Q plot now confirms that the normality assumption is largely satisfied for active students. This suggests our linear model is quite effective at predicting performance *given* that a student participates in the exam.

2.  **The "Absences" Flip**: In the full model (including zeros), absences appeared to have a positive or negligible coefficient. In the refined model, the coefficient flips to **negative ($\approx -0.055$)**. This proves the 0-grade noise was masking the true negative impact of missing class: once we remove dropouts, we see that for active students, more absences lead to lower grades.

3.  **Selection Bias in School Support**: The strong negative coefficient for school support ($\beta \approx -2.09$, $p < 0.001$) suggests that this variable acts as a **proxy for struggling students**. It is not that support *causes* lower grades, but rather that support is given primarily to students already at risk of lower performance.

**Final Summary**:
Linear Regression works well for **scoring** (predicting the exact grade of a regularly attending student), driven by *Academic History (Failures)*, *Family Background (Medu)*, and *Behavior (Absences)*. However, the event of **administrative failure (Grade 0)** is a structural break that cannot be modeled continuously.
