---
title: "Final Project"
subtitle: "Inferential Statistics 2025/2026"
author: "Giordano Alberti, Leonardo Bertani"
date: "`r Sys.Date()`"
output: 
  pdf_document: 
    toc: true
    fig_height: 6
    number_sections: true
  html_document: 
    toc: true
editor_options: 
  markdown: 
    wrap: 80
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\newpage

# Abstract

This report analyzes the determinants of academic performance using the UCI "Student Performance Data Set," derived from Mathematics classes in two Portuguese schools (Gabriel Pereira and Mousinho da Silveira) and collected through school reports and questionnaires. The study aims to identify the demographic, social, and behavioral factors driving final grades (G3), using empirical evidence to validate or refuse common assumptions about student performance. Employing Chi-Squared tests, along with parametric tests and regression analysis, we investigate variables ranging from family background to social habits.

# Introduction

```{r}
data <- read.csv("./data/student-mat.csv", sep=";", stringsAsFactors = TRUE)
str(data)
summary(data)
head(data)
```


# Statistical analysis

## Test of Correlation

While the Chi-Squared, or in some cases the Fisher's Exact Test, allowed us to analyze associations between categorical variables, to examine relationship between numerical variables we use the Correlation Analysis.

This test calculate a coefficient that measures the strength and the direction of the relationship between the two variables. The most common is the Pearson coefficient ($r$), but depending on the data distribution, the Spearman rank coefficient ($\rho$) may be more appropriate.

The coefficient for Pearson's test is: $$r = \frac{1}{n-1}\sum_{i=1}^n \left(\frac{x_i - \bar{x}}{s_x}\right) \left(\frac{y_i-\bar{y}}{s_y}\right)$$ where $\bar{x}$ and $\bar{y}$ are the sample means, $s_x$ and $s_y$ are the sample standard deviation of $x$ and $y$. 
The coefficient for Spearman's test is: $$\rho = \frac{cov(R[X],R[Y])}{\sigma_{R[X]} \sigma_{R[Y]}}$$
the correlation operator applied to the rank variables.

Then, this value is inside the range $[-1,+1]$:
\begin{itemize}
    \item $r, \rho > 0$: positive correlation
    \item $r, \rho < 0$: negative correlation
    \item $r, \rho = 0$: no correlation.
\end{itemize}

The tes test verifies the following hypotheses:
\begin{itemize}
    \item $H_0$: variables are not correlated ($r \lor \rho = 0$)
    \item $H_1$: there is a correlation between variables ($r \lor \rho \neq 0$).
\end{itemize}

### Impact of Absences on Final Grade

We hypothesize a negative relationship between the number of school absences (`absences`) and the final grade (`G3`). A high number of absences could affect ability to keep up with the study program and lowering performance.

To choose the most appropriate correlation test, we have to verify the distribution of the `absences` variable. In fact, the Pearson Correlation test assumes that data are drawn from a normal distribution.

```{r echo=FALSE}
hist(data$absences, main="Distribution of Absences", xlab="Number of Absences")
shapiro_test <- shapiro.test(data$absences)
```

As shown in the histogram, the variable follows a right-skewed distribution (strongly asymmetric). 

Then, the Shapiro-Wilk's test returns a p-value = `r shapiro_test$p.value` lower than $\alpha = 0.05$ that confirms non-normality. 
Since those information, we can say that is better to compute a Spearman test which not required a normal distribution of the data and is more resistent to outliers.

```{r include=FALSE}
cor_spearman_absences <- cor.test(data$absences, data$G3, method = "spearman", exact = FALSE)
```

After computing the test, we can see that the correlation factor from Spearman is `r round(cor_spearman_absences$estimate, 3)`.

Due to the presence of ties in the `absences` and `G3` variables, the exact p-value for Spearman's $\rho$ could not be computed. Therefore, the standard asymptotic approximation was used `r round(cor_spearman_absences$p.value, 3)`. Since this value is very hight, we cannot reject $H_0$. Statistically, this indicates no significant correlation between absences and the final grade in this dataset. This finding refutes our initial hypothesis that higher absences lead to lower grades.

Using a scatterplot we can see that the trend given by the correlation test.

```{r echo=FALSE}
plot(data$absences, data$G3, 
     main = "Absences vs Final Grade (G3)",
     xlab = "Number of Absences", 
     ylab = "Final Grade (G3)", 
     pch = 19, col = rgb(0,0,1,0.5))
abline(lm(data$G3 ~ data$absences), col = "red")
```