---
title: "Final Project"
subtitle: "Inferential Statistics 2025/2026"
author: "Giordano Alberti, Leonardo Bertani"
date: "`r Sys.Date()`"
output: 
  pdf_document: 
    toc: true
    fig_height: 6
    number_sections: true
  html_document: 
    toc: true
editor_options: 
  markdown: 
    wrap: 80
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\newpage

# Abstract

This report analyzes the determinants of academic performance using the UCI "Student Performance Data Set," derived from Mathematics classes in two Portuguese schools (Gabriel Pereira and Mousinho da Silveira) and collected through school reports and questionnaires. The study aims to identify the demographic, social, and behavioral factors driving final grades (G3), using empirical evidence to validate or refuse common assumptions about student performance. Employing Chi-Squared tests, along with parametric tests and regression analysis, we investigate variables ranging from family background to social habits.

# Introduction

```{r include=FALSE}
data <- read.csv("./data/student-mat.csv", sep=";", stringsAsFactors = TRUE)
str(data)
summary(data)
head(data)
```


## Logistic Regression

Logistic Regression is the standard statistical method for binary classification problems.
It models the probability $\vartheta(X)=P(Y=1|X)$ that an event occurs as a function of independent variables $X$. Instead of fitting a straight line, it uses the logistic function to map the output between 0 and 1.

The logistic function is defined as:
$$\vartheta(X) = \frac{e^{\beta_0 + \beta_1 X_1 + \dots + \beta_k X_k}}{1 + e^{\beta_0 + \beta_1 X_1 + \dots + \beta_k X_k}}$$
and the regression is expressed in terms of log-odds: $$\log\left(\frac{\vartheta(X)}{1-\vartheta(X)}\right) = \beta_0 + \beta_1 X_1 + \dots + \beta_k X_k$$
Where $\beta_i$ are the coefficients of the model.

A positive coefficient ($\beta_i > 0$) implies that as $X_i$ increases, the odds/probability of success increase. Conversely, a negative coefficient suggests a decrease in the probability of success.

### The Impact of Educational Support

We aim to predict whether a student will be Promoted (G3 $\ge$ 10) or Fail (G3 $<$ 10) according to Portuguese grading system.

#### Initial predictors and multicollinearity

The initial predictors selected, from a strictly academic point of view, are:
\begin{itemize}
    \item \texttt{schoolsup:} Extra educational support [Y/N].
    \item \texttt{paid:} Extra paid classes within the course subject [Y/N].
    \item \texttt{higher:} Wants to take higher education [Y/N].
\end{itemize}

Before computing the model, we check for multicollinearity among predictors (`schoolsup`, `paid`, `higher`). High correlation between predictors can inflate the variance of the coefficient estimates, making the model unstable. 

```{r echo = FALSE}
library(corrplot)
binary_vars <- data[, c("schoolsup", "paid", "higher")]

# cor accept only numeric variables: if value is "yes" -> 1, else -> 0
binary_vars_num <- as.data.frame(lapply(binary_vars, function(x) ifelse(x == "yes", 1, 0)))
cor_matrix <- cor(binary_vars_num, use = "complete.obs")
corrplot(cor_matrix, method = "color", type = "upper", tl.cex = 0.7, addCoef.col = "black",title = "Correlation Matrix", number.cex = 0.6)
```

The plot shows low correlation for `schoolsup` vs `paid` and `schoolsup` vs `higher` (values close to 0). For `paid` vs `higher`, there is a moderate positive correlation (`r round(cor_matrix["paid", "higher"],2)`), confirming that students aiming for university are more likely to invest in paid classes. However, since the value is below critical thresholds ($ < 0.7$), we proceed with the regression.

#### Models evaluation

##### Initial selection

```{r include=FALSE}
data$Promoted <- ifelse(data$G3 >= 10, 1, 0)
logit_model <- glm(Promoted ~ schoolsup + paid + higher, 
                   data = data, family = binomial)
logit_model_summary <- summary(logit_model)
```

The initial model reveals that higher is a significant positive predictor (`r round(logit_model_summary$coefficients["higheryes", "Pr(>|z|)"],4)` $< 0.05$), while `schoolsup` has a negative coefficient. This highlights the "support paradox": educational support is often a proxy for academic difficulty rather than a direct booster of success in the short term.

```{r include=FALSE}
logit_improved <- glm(Promoted ~ schoolsup + paid + higher + failures + goout, 
                      data = data, family = binomial)

logit_improved_summary <- summary(logit_improved)
```

To improve the model's accuracy (with lower AIC), we introduce two variables related to background and social activity:
\begin{itemize} 
  \item \texttt{failures:} Number of past class failures. 
  \item \texttt{goout:} Frequency of going out with friends. 
\end{itemize}

Now, the statistically significant coefficients are, in ascending order, `schoolsup`, `goout` and `failures`.

##### Validation via Stepwise Selection

To validate our variable selection, we applied an automated stepwise selection algorithm (AIC-based) on an extended set of predictors.

```{r include=FALSE}
logit_full <- glm(Promoted ~ schoolsup + paid + higher + failures + goout + 
                            absences + studytime + activities + nursery + internet + 
                            address + Pstatus + famrel + freetime + health + sex + Dalc + Walc, 
                            data = data, family = binomial)
logit_stepwise <- step(logit_full, direction = "both", trace = 0)
logit_stepwise_summary <- summary(logit_stepwise)
```

The algorithm converged to a model containing: `r paste(names(coef(logit_stepwise))[-1], collapse = ", ")`.

Comparing the AIC scores:
\begin{itemize} 
  \item Manual Initial Model AIC: `r round(logit_model$aic,0)`
  \item Manual Improved Model AIC: `r round(logit_improved$aic, 0)`
  \item Stepwise Model AIC: `r round(logit_stepwise$aic, 0)` \end{itemize}
  
#### Data Refinement: Excluding Zero-Grades

We observed that approximately 38 entries have a final grade (`G3`) of 0. In this context, a grade of 0 likely indicates students who were absent or dropped out, rather than a genuine low-performance assessment.
We tested the "Improved Model" structure on a filtered dataset to assess performance on active students.

```{r include=FALSE}
data_filtered <- subset(data, G3 != 0)
data_filtered$Promoted <- ifelse(data_filtered$G3 >= 10, 1, 0)
logit_filtered <- glm(Promoted ~ schoolsup + failures + goout + paid + higher, 
                      data = data_filtered, family = binomial)
logit_filtered_summary <- summary(logit_filtered)
```

#### Comparison between models

We compare the discriminative ability of all four models using the ROC Curve.

```{r include=FALSE}
library(pROC)

prob_model <- predict(logit_model, type = "response")
prob_improved <- predict(logit_improved, type = "response")
prob_stepwise <- predict(logit_stepwise, type = "response")
prob_filtered <- predict(logit_filtered, type = "response")

roc_model <- roc(data$Promoted, prob_model)
roc_improved <- roc(data$Promoted, prob_improved)
roc_stepwise <- roc(data$Promoted, prob_stepwise)
roc_filtered <- roc(data_filtered$Promoted, prob_filtered)
```
```{r message=FALSE, echo=FALSE}
plot(roc_model, col="red", main="ROC Curve Comparison", lwd=2)
plot(roc_improved, col="blue", add=TRUE, lwd=2)
plot(roc_stepwise, col="green", add=TRUE, lwd=2)
plot(roc_filtered, col="magenta", add=TRUE, lwd=2)
legend("bottomright", 
       legend=c(paste("Initial Model (AUC =", round(auc(roc_model), 2), ")"), 
       paste("Improved Model (AUC =", round(auc(roc_improved), 2), ")"),
       paste("Stepwise Model (AUC =", round(auc(roc_stepwise), 2), ")"),
       paste("Filtered Model (AUC =", round(auc(roc_filtered), 2), ")")),
       col=c("red", "blue", "green", "magenta"), lwd=2)
```

Visually, the graph illustrates a clear shift of the Improved Model's curve toward the top-left corner compared to the Initial Model, confirming a substantial gain in predictive accuracy. The performance of the Improved Model and the Stepwise model are widely consistent.
Furthermore, the performance of the Filtered Model (AUC $\approx$ `r round(auc(roc_filtered), 2)`) remains highly comparable to the Improved Model (AUC $\approx$ `r round(auc(roc_improved), 2)`). This stability is a crucial finding: it demonstrates that our model is robust and does not rely solely on the easy classification of dropouts (`G3` $=0$) to achieve good results. It effectively distinguishes between passing and failing students.
