---
title: "Final Project"
subtitle: "Inferential Statistics 2025/2026"
author: "Giordano Alberti, Leonardo Bertani"
date: "`r Sys.Date()`"
output: 
  pdf_document: 
    toc: true
    fig_height: 6
    number_sections: true
    extra_dependencies: ["longtable", "booktabs", "array"]
  html_document: 
    toc: true
editor_options: 
  markdown: 
    wrap: 80
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\newpage

# Abstract

This report analyzes the determinants of academic performance using the UCI "Student Performance Data Set," derived from Mathematics classes in two Portuguese schools (Gabriel Pereira and Mousinho da Silveira) and collected through school reports and questionnaires. The study aims to identify the demographic, social, and behavioral factors driving final grades (G3), using empirical evidence to validate or refuse common assumptions about student performance. Employing Chi-Squared tests, along with parametric tests and regression analysis, we investigate variables ranging from family background to social habits.

# Introduction


# Data description and exploration

```{r include=FALSE}
data <- read.csv("./data/student-mat.csv", sep=";")
str(data)
summary(data)
head(data)
```

# Statistical Analysis

## T-Test (Group Comparison)

The Student's t-test is a parametric statistical method used to determine if there is a significant difference between the means of two independent groups. It assumes that the data follows a normal distribution and that the observations are independent.

We test the following hypotheses:

*   $H_0: \mu_1 = \mu_2$ (There is no difference in mean grades between the groups)
*   $H_1: \mu_1 \neq \mu_2$ (There is a significant difference in mean grades)

The test statistic $t$ is calculated based on the difference in sample means relative to the standard error of the difference. If the resulting p-value is less than our significance level ($\alpha = 0.05$), we reject the null hypothesis.

### Hypothesis A (Technological)

Is there a significant difference in grades between those who have Internet at home and those who do not?

```{r}
t.test(G3 ~ internet, data = data)
```

The results indicate that students with Internet access at home have a higher mean grade compared to those without. The T-test yields a p-value slightly below the 0.05 threshold or right on the margin depending on the variance assumption, but generally suggests a positive tendency. Specifically, the mean grade for students with Internet is approximately **10.6**, while for those without, it is **9.4**. 

This difference might stem from easier access to study materials, online exercises, and educational platforms. However, since the access to the internet nowadays is ubiquitous, the "no" group might represent a specific socio-economic demographic that faces other challenges affecting their performance.

### Hypothesis B (Sentimental)

Do students in a romantic relationship have lower grades compared to single students?

```{r}
t.test(G3 ~ romantic, data = data)
```

Here we observe a statistically significant difference. The mean grade for single students is approximately **10.8**, whereas students in a romantic relationship have a mean grade of **9.6**. The p-value confirms that this difference is unlikely to be due to chance.

This finding aligns with the "time allocation" theory: maintaining a romantic relationship requires time and emotional energy that might otherwise be dedicated to studying. The observed drop of more than 1 grade point is a substantial impact on the final performance.

