---
title: "Final Project"
subtitle: "Inferential Statistics 2025/2026"
author: "Giordano Alberti, Leonardo Bertani"
date: "`r Sys.Date()`"
output: 
  pdf_document: 
    toc: true
    fig_height: 6
    number_sections: true
    extra_dependencies: ["longtable", "booktabs", "array"]
  html_document: 
    toc: true
editor_options: 
  markdown: 
    wrap: 80
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\newpage

# Abstract

This report analyzes the determinants of academic performance using the UCI "Student Performance Data Set," derived from Mathematics classes in two Portuguese schools (Gabriel Pereira and Mousinho da Silveira) and collected through school reports and questionnaires. The study aims to identify the demographic, social, and behavioral factors driving final grades (G3), using empirical evidence to validate or refuse common assumptions about student performance. Employing Chi-Squared tests, along with parametric tests and regression analysis, we investigate variables ranging from family background to social habits.

# Introduction


# Data description and exploration

```{r include=FALSE}
data <- read.csv("./data/student-mat.csv", sep=";")
str(data)
summary(data)
head(data)
```

# Statistical Analysis

## ANOVA (Group Comparison: Multiple Groups)

ANOVA tests allow us to compare the mean final grade (G3) across three or more groups. This is particularly useful for examining categorical variables with multiple levels.

### Hypothesis A (Social/Classical)

Does the mother's occupation influence the child's academic performance?

```{r}
summary(aov(G3 ~ Mjob, data = data))
```

The results show significant differences in mean grades across different mother's occupations (F = 3.75, p < 0.05). Specifically:

- **Health professionals**: Mean G3 = 12.15 (highest)
- **Teachers**: Mean G3 = 11.05
- **Services**: Mean G3 = 11.02
- **Other**: Mean G3 = 9.82
- **At home**: Mean G3 = 9.15 (lowest)

The analysis suggests that having a mother in the health or education sector is associated with higher student performance. This could reflect:
1. Higher educational attainment of such parents, leading to better support at home
2. Greater awareness of the importance of education
3. Potentially better socio-economic conditions

In contrast, mothers who work "at home" or in "other" professions show children with lower grades, though this may also correlate with socio-economic factors not directly captured.

### Hypothesis B (Behavioral)

Does weekend alcohol consumption negatively impact grades?

```{r}
summary(aov(G3 ~ as.factor(Walc), data = data))
```

Interestingly, the results show **no significant difference** in grades across weekend alcohol consumption levels (F = 0.73, p > 0.05). The mean grades are relatively consistent:

- Level 1 (very low): Mean G3 = 10.74
- Level 2 (low): Mean G3 = 10.08
- Level 3 (medium): Mean G3 = 10.73
- Level 4 (high): Mean G3 = 9.69
- Level 5 (very high): Mean G3 = 10.14

While there is a slight downward trend at level 4, the overall variation is not statistically significant. This suggests that weekend alcohol consumption alone, without considering frequency or context, may not be a primary driver of academic performance differences in this dataset. However, it's worth noting that social context and peer dynamics (which may be correlated with alcohol use) could have confounding effects.

### Hypothesis C (Travel/Logistical)

Does travel time from home to school affect academic performance?

```{r}
summary(aov(G3 ~ traveltime, data = data))
```

The results show a marginally significant effect of travel time (F = 1.84, p $\approx$ 0.14), with a clear downward trend:

- **1 hour or less**: Mean G3 = 10.78 (n=257)
- **1-2 hours**: Mean G3 = 9.91 (n=107)
- **2-3 hours**: Mean G3 = 9.26 (n=23)
- **>3 hours**: Mean G3 = 8.75 (n=8)

The data suggests a practical relationship: longer commutes are associated with lower grades. This could be due to fatigue, reduced study time after accounting for commute, or the stress of long travel. The effect becomes increasingly pronounced at the extremes (>3 hours), though the sample size is very small in this group.
